Making Sequential Programs Deterministic

### 2019-04-06 randoop-case-study.tex
Start adding case study comments

To evaluate the usability of \theDeterminismChecker,
we applied it to the Randoop test
generator~\ifanonymous\citepalias{randoop-tool}\else\cite{PachecoLEB2007}\fi.
We also wrote specifications for libraries Randoop uses, such as the JDK, JUnit,
and others.
Like \theDeterminismCheckerImplementation,
all the case study materials are publicly available at \myurl{\todo{url for
    case study}} for reproducibility.

@@

The developers took extensive action to detect and mitigate the effect of nondeterminism.
They used Docker images to run tests, to avoid system dependencies such as
a different JDK having a different number of classes or methods.
They wrote tests with relaxed oracles (assertions) that permit multiple
possible answers --- for example, in code coverage of generated tests.
They used linters such as Error Prone, which warns if \<toString> is used on
array objects, which don't override \<Object.toString> and therefore print a
hash code which may vary from run to run.
They used a library that makes hash codes deterministic, by giving each
object of a type a unique ID that counts up from 1 rather than using a
memory address as \<Object.hashCode> does.
They wrote specialized tools to preprocess output and logs to make them easier
to compare, such as by removing or canonicalizing hash codes, dates, and
other nondeterministing output (commits e15f9155--32f72234).
Each of these efforts was helpful, but they were not enough to solve the problem.

In July 2017, the Randoop developers spent two weeks of full-time work to
eliminate unintentional nondeterministic behavior in Randoop.
Their methodology was to repeatedly
 run Randoop with verbose logging enabled,
 look for differences in logging output,
 find the root cause of nondeterminism,
 and eliminate it.
Some of the nondeterminism was in libraries, such as the JDK\@.
The most common causes were \<toString> routines and iteration order of sets and maps.
The most common fixes were to change the implementation of \<toString>
and to use \<LinkedHashSet> and \<LinkedHashMap> or to sort collections
before iterating over them.
The developers did not make every \<Set> and \<Map> a \<LinkedHashSet> or \<LinkedHashMap>, because that
would have increased memory and CPU costs.  They chose not to make every
order-nondeterministic \<List> a \<Set>, for similar reasons:  deduplication was
not always desired, and even where it was acceptable, it would have increased costs.

@@

In brief, we wrote type qualifiers in the Randoop source code to express its
determinism specification
We then ran
\theDeterminismChecker.  Each warning indicated a mismatch between the
specification and the implementation, which we addressed by changing our
specification or reporting a bug in the implementation.

@@

We annotated one package at a time, starting with the most central packages
that are most depended upon.  Within a package, we followed a similar
strategy, annotating supertypes first.  If the determinism of classes and methods had been
documented, then our annotation effort
would have been easy, just converting English into type qualifiers.
Unfortunately, this was not the case: we had to reverse-engineer each
specification, largely from the methods it calls.  If a method $m$ calls a
method that we had previously annotated as nondeterministic, we annotated
$m$ as nondeterministic.
Sometimes after we
annotated a subclass, we had to go back and change existing annotation on
the supertype or elsewhere because the subclass implementation was
inconsistent with the specification we had guessed for the supertype.
After resolving most type-checking warnings by adding or changing
annotations, we manually examined each remaining one to determine whether
it was a bug or a false positive warning.

@@

\item[Moderate issues]
  \strut
\begin{description}
\item[\bugHashMapOutput]
While printing diagnostic output, Randoop iterated over a \<HashMap> in
arbitrary order, making the output difficult to compare across different
executions, or even to find information in.
The class already implemented \<Comparable>, so
the developers changed \<methodWeights.keySet()> to
\codeid{new TreeSet<>(methodWeights.keySet())} in a \<for> loop (commit
f212cc7e).
% This was actually introduced after the coding sprint.  Maybe some other
% defects were too, so our criticism should be about regressions rather
% than not finding them in the first place.
\end{description}

@@

\item[\bugHashcodeOutput]
  Diagnostic output printed a hash code for brevity.
  The developers changed it to have deterministic output (commit 661a4970).
  Although the Randoop developers classified this issue as minor, it is
  similar to ones they fixed during their two-week coding sprint.

\item[\bugTimestampOutput]
  Diagnostic output printed a timestamp.
  The developers fixed it by making it obey an
  option about whether to print timestamps (commit a460df97).
  The developers may not have noticed this because their log-postprocessing
  tools removed them from the log when doing comparisons.

@@

\TheDeterminismChecker issued 37 false positive warnings, or 1 every 650
lines of code.

The most common reasons (responsible for 67\% of false positive warnings)
were:

\begin{itemize}
\item   (22\%)
%      8 constructors guarantee all instances of this class are @Det
  When every instance of a class is \<@Det>, \<@PolyDet> is equivalent to
  \<@Det>, but \theDeterminismChecker does not yet recognize this fact.
\item  (16\%)
%      6 https://github.com/t-rasmud/checker-framework/issues/96
  For a type that cannot be \<@OrderNonDet>, \<@PolyDet("up")> is
  equivalent to \<@PolyDet>.
\item (14\%)
%      5 can't specify receiver for anonymous class
  Java does not have syntax for specifying the type of an anonymous class.
\item (11\%)
%      4 actual type toString returns @PolyDet
  A variable is declared as \<Object> which has nondeterministic \<toString>,
  but at run time the value always has deterministic \<toString>.  For
  example, the run-time value is always an \<Integer> or a \<String>.
\item (5\%)
%      2 iterating over @OrderNonDet collection to modify another
  Iteration over an \<@OrderNonDet> collection in order to make another
  \<@OrderNonDet> collection, but the extracted elements have type
  \<@NonDet>.  For example, the following code is safe, but
  \theDeterminismChecker issues a warning.  Variable \<elt> has type
  \<@NonDet>, so the call to \<add> does not typecheck.
\begin{Verbatim}
  void m(@OrderNonDet List<@Det String> input) {
    @OrderNonDet List<@Det String> output = new List<>();
    for (String elt : input) {  // elt has type @NonDet String
      output.add(elt);          // error: illegal argument
    }
  }
\end{Verbatim}

\end{itemize}

The first two issues can be resolved by improvements to
\theDeterminismCheckerImplementation.
The third and fourth issues are limitations of the Java language, but by a
local refactoring in the Randoop codebase can make the code typecheck.
Refactoring can also eliminate the last warning.

@@

One is that we annotated some core components of Randoop to require all
Boolean expressions used in conditionals to be deterministic.  (This is an
option in \theDeterminismCheckerImplementation, which increases soundness;
see \cref{sec:threats}.)  This forced
their clients --- transitively, much of the Randoop codebase --- to be
annotated as deterministic.
These annotations are not incorrect, but they are stricter than necessary
because some method types could otherwise be qualifier-polymorphic.
This would make them usable in more contexts.  It would also reduce the
number of necessary annotations, since qualifier-polymorphism (the
\<@PolyDet> annotation) is the default for
formal parameter and return types.  We plan to re-do the annotation effort
from scratch, and the number of annotations will be much lower.

@@

% LocalWords:  linters preprocess canonicalizing nondeterministing bdeebd
% LocalWords:  e15f9155 b473fd14 deduplication c15ccbf2 ff5b4c4 eda87f src
% LocalWords:  f72234 a8e63fb c975a9f7 e3c56 a4970 a460df97 postprocessing
% LocalWords:  MinimizerTests SpecificationCollection findOverridden Det

>>>>

To evaluate the usability of \theDeterminismChecker,
we applied it to the Randoop test
generation tool~\ifanonymous\citepalias{randoop-tool}\else\cite{PachecoLEB2007}\fi.
We also wrote specifications for libraries it uses, such as the JDK, JUnit,
and others.
Like \theDeterminismCheckerImplementation,
all the case study materials are publicly available at \myurl{\todo{url for
    case study}} for reproducibility. \todo{This should include our link to
      our annotated plume-util, which we found was required to reproduce the
    results of the case study}

@@

The developers took extensive action to detect and mitigate the effect of nondeterminism.
They used Docker images to run tests, to avoid system dependencies such as
a different JDK having a different number of classes or methods.
They wrote tests with relaxed oracles (assertions) that permit multiple
possible answers --- for example, in code coverage of generated tests.
They used linters such as Error Prone, which warns if \<toString> is used on
array objects, which don't override \<Object.toString> and therefore print a
hash code which may vary from run to run.
They used a library that makes hash codes deterministic, by giving each
object of a type a unique ID that counts up from 1 rather than using a
memory address as \<Object.hashCode> does.
They wrote specialized tools to preprocess output and logs to make them easier
to compare, such as by removing or canonicalizing hash codes, dates, and
other nondeterministing output (commits e15f9155--b473fd14).
Each of these efforts was helpful, but they were not enough to solve the problem.

In July 2017, the Randoop developers spent two weeks of full-time work to
eliminate unintentional nondeterministic behavior in Randoop.
Their methodology was to repeatedly
 run Randoop with verbose logging enabled,
 look for differences in logging output,
 find the root cause of nondeterminism,
 and eliminate it.
Some of the nondeterminism was in libraries, such as the JDK\@.
The most common causes were \<toString> routines and iteration order of sets and maps.
The most common fixes were to change the implementations of \<toString>
and to use \<LinkedHashSet> and \<LinkedHashMap> or to sort collections
before iterating over them.
The developers did not make every \<Set> and \<Map> a \<LinkedHashSet> or \<LinkedHashMap>, because that
would have increased memory and CPU costs.  They chose not to make every
order-nondeterministic \<List> a \<Set>, for similar reasons:  deduplication was
not always needed and would have increased costs.

@@

In brief, we wrote type qualifiers in the Randoop source code to express its
determinism specification.
We then ran
\theDeterminismChecker.  Each warning indicated a mismatch between the
specification and the implementation, which we addressed by changing our
specification or reporting a bug in the implementation.

@@

We annotated one package at a time, starting with the most central packges
that are most depended upon.  Within a package, we followed a similar
strategy, annotating supertypes first.  If the determinism of classes and methods had been
documented, then our annotation effort
would have been easy, just converting English into type qualifiers.
Unfortunately, this was not the case: we had to reverse-engineer each
specification, largely from the methods it calls.  If a method $m$ calls a
method that we had previously annotated as nondeterministic, we annotated
$m$ as nondeterministic.
Sometimes after we
annotated a subclass, we had to go back and change existing annotation on
the supertype or elsewhere because the subclass implementation was
inconsistent with the specification we had guessed for the supertype.
After resolving most type-checking warnings by adding or changing
annotations, we manually examined each remaining one to determine whether
it was a bug or a false positive warning.

@@

@@

\item \bugHashcodeOutput:
  Diagnostic output printed a hash code for brevity.
  The developers changed it to have deterministic output (commit 661a4970).
  Although the Randoop developers classified this issue as minor, it is
  very similar to ones they fixed during their two-week coding sprint.

\item \bugTimestampOutput:
  Diagnostic output printed a timestamp.
  The developers fixed it by making it obey an
  option about whether to print timestamps (commit a460df97).
  The developers may not have noticed this because their log-postprocessing
  tools removed them from the log when doing comparisons.

@@

\todo{Java 8 lambdas could help.}

@@

One is that we annotated some core components of Randoop to require all
boolean expressions used in conditionals to be deterministic.  (This is an
option in \theDeterminismCheckerImplementation, which increases soundness;
see \cref{sec:threats}.)  This forced
their clients --- transitively, much of the Randoop codebase --- to be
annotated as deterministic.
These annotations are not incorrect, but they are stricter than necessary
because some method types could otherwise be qualifier-polymorphic.
This would make them usable in more contexts.  It would also reduce the
number of necessary annotations, since qualifier-polymorphism (the
\<@PolyDet> annotation) is the default for
formal parameter and return types.  We plan to re-do the annotation effort
from scratch, and the number of annotations will be much lower.

@@
